---
title: "PCA & MDS  for Country  GDP and Life Expectancy"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library_insert, echo=FALSE}
#import library
library(dplyr)
library(ggplot2)
library(psych)
library(rgl)

```








read data
```{r}
#read data
dat <- read.csv('gap.csv')

#get gdp data
gdp <- dat[,3:14]
years <- seq(1952, 2007,5)
colnames(gdp) <- years
rownames(gdp) <- dat[,2]

#get lifeExp
lifeExp <- dat[,15:26]
colnames(lifeExp) <- years
rownames(lifeExp) <- dat[,2]

```









explore how GDP and life expectancy have changed


trans data to be structural
```{r}
#structure the gdp data for plot
gdp_temp <- data.frame()
gdp_temp <- data.frame(gdpPercap=unlist(gdp))
gdp_temp$year <- as.integer(substring(rownames(gdp_temp),1,4))
gdp_temp$continent <- dat[,1]
gdp_temp$country <- dat[,2]
rownames(gdp_temp) <- NULL

gdp_struc <- gdp_temp

#structure the lifeExp data for plot
lifeExp_temp <- data.frame()
lifeExp_temp <- data.frame(lifeExp=unlist(lifeExp))
lifeExp_temp$year <- as.integer(substring(rownames(lifeExp_temp),1,4))
lifeExp_temp$continent <- dat[,1]
lifeExp_temp$country <- dat[,2]
rownames(lifeExp_temp) <- NULL

lifeExp_struc <- lifeExp_temp
```



plot trend of GDP and life expectancy
```{r}
#calculate mean gdp for overall trend
gdp_struc_mean <- 
        gdp_struc %>%
        group_by(year) %>%
        summarise(n=n()
                  ,gdpPercap_mean = mean(gdpPercap))
#plot overall gdp trend
g <- ggplot(gdp_struc_mean, aes(year, gdpPercap_mean))
g +
        geom_point(lwd=3,col='steelblue') +
        geom_line(lwd=1.2,col='steelblue') +
        theme_bw()
#calculate mean lifeExp for overall trend
lifeExp_struc_mean <- 
        lifeExp_struc %>%
        group_by(year) %>%
        summarise(n=n()
                  ,lifeExp_mean = mean(lifeExp))
#plot overall lifeExp trend
g <- ggplot(lifeExp_struc_mean, aes(year, lifeExp_mean))
g +
        geom_point(lwd=3,col='steelblue') +
        geom_line(lwd=1.2,col='steelblue') +
        theme_bw()
```
Overall, both the GDP Per ceptia and the life expectancy growth during 1952 and 2007. From 1972 to 1992, the growth rate is lower.



calculate the average life expectancy and GDP per capita for each continent through time and plot these
```{r}
#calculate mean gdp for trend by continent
gdp_struc_mean <- 
        gdp_struc %>%
        group_by(continent,year) %>%
        summarise(n=n()
                  ,gdpPercap_mean = mean(gdpPercap))
#plot gdp trend by continent
g <- ggplot(gdp_struc_mean, aes(year, gdpPercap_mean, color = continent))
g +
        geom_point() +
        geom_line() +
        theme_bw()
#calculate mean lifeExp for trend by continent
lifeExp_struc_mean <- 
        lifeExp_struc %>%
        group_by(continent,year) %>%
        summarise(n=n()
                  ,lifeExp_mean = mean(lifeExp))
#plot lifeExp trend by continent
g <- ggplot(lifeExp_struc_mean, aes(year, lifeExp_mean, color = continent))
g +
        geom_point() +
        geom_line() +
        theme_bw()
```
Countries in Europe, Americas and Oceania has higher GDP per capita and higher life expectancy. From 1972 to 1997, GDP per captia of Asia, Americas and Africa growed slowly. From 1987 to 2002, life expectancy of Africa growed slowly.








gdp principal component analysis
```{r}
#gdp
#work with the log of the GDP
log_gdp <- log(gdp)
#standarization
log_gdp_scale = scale(log_gdp)
#calculate corr
log_gdp_cor <- cor(log_gdp_scale)
#Calculate eigenvalues and eigenvectors
log_gdp_eigen=eigen(log_gdp_cor)
log_gdp_var=log_gdp_eigen$values
options(digits=4, scipen=4)
log_gdp_var
log_gdp_pca_var=log_gdp_var/sum(log_gdp_var)
log_gdp_pca_var
#scree plot
g <- ggplot(,aes(x=1:12,y=log_gdp_pca_var))
g+geom_point(lwd=3,col='steelblue')+
        geom_line(col='steelblue',lwd=1.2)+
        labs(x = "Principal Component", y = "Proportion of Variance Explained")+
        theme_bw()
```
For log_gdp,first components stand for vast majority of the variance. The proportion of variance Explained can be presented in the scree graph. Already two first components stand for almost 98% of variance. Each next component gives not more than 1%.
Thus, retaining the first two components is a good choice.



lifeExp principal component analysis
```{r}
#lifeExp
#standarization
lifeExp_scale = scale(lifeExp)
#calculate corr
lifeExp_cor <- cor(lifeExp_scale)
#Calculate eigenvalues and eigenvectors
lifeExp_eigen=eigen(lifeExp_cor)
lifeExp_var=lifeExp_eigen$values
options(digits=4, scipen=4)
lifeExp_var
lifeExp_pca_var=lifeExp_var/sum(lifeExp_var)
lifeExp_pca_var
#scree plot
g <- ggplot(,aes(x=1:12,y=lifeExp_pca_var))
g+geom_point(lwd=3,col='steelblue')+
        geom_line(col='steelblue',lwd=1.2)+
        labs(x = "Principal Component", y = "Proportion of Variance Explained")+
        theme_bw()
```
For lifeExp, first components stand for vast majority of the variance. The proportion of variance Explained can be presented in the scree graph. Already two first components stand for almost 98% of variance. Each next component gives not more than 2%. 
Thus, retaining the first two components is a good choice.









scatter plots of combinations of the first three principal component scores
```{r}
#merge lifeExp and gdp
combined <- cbind(log(dat[,3:14]), dat[,15:26])
#PCA analysis
combined_pca = princomp(combined,cor=T)
#summary(combined_pca)
#Extract the first three components
combined_3D = combined_pca$scores[,1:3]
colnames(combined_3D) <- c('PC1','PC2','PC3')
combined_3D <- data.frame(combined_3D)
#label preparation for plot
continent <- dat[,1]
country <- dat[,2]
#plot of the new 3 PC df
#3D plot for exploration
plot3d(combined_3D,col=as.numeric(dat[,1]),radius=50)
#2D plot for explaination
g <- ggplot(combined_3D,aes(x=PC1,y=PC2,color=continent))
g+geom_point(lwd=2)+
        geom_text(aes(label = country),nudge_y = 0.25)+
        theme_bw()
```
Countries in Europe, Americas and Oceania have higher value in PC1. While Africa countries have small value in PC2(most of them is smaller than 0, which means that they have lower value then the average). To explain it, let's make a hypothesis that this is because countries in Europe, Americas and Oceania has higher GDP per capita and higher life expectancy. This can also be shown in the visualization of the lifeExp and gdpPercap by continent.










Perform multidimensional scaling
plot a 2-dimensional representation of the data
```{r}
#MDS
#calculate distance
combined.dist <- dist(combined, method = "euclidean")
#MDS analysis
combined.MDS <- cmdscale(combined.dist, k = 2)
summary(combined.MDS)
#make the MDS result a df
colnames(combined.MDS) <- c('V1','V2')
combined.MDS <- data.frame(combined.MDS)
#MDS plot
g <- ggplot(combined.MDS,aes(x=V1,y=V2,color=continent))
g+
        geom_point(lwd=2)+
        geom_vline(aes(xintercept=0),colour="#BB0000", linetype="dashed")+
        theme_bw()
```
All the points are transformed to new coordinate 2D plane named V1, V2, respectively. The points of Africa countries are on the left side of the vertical line(V1 = 0), and the points of Europe, Americas and Oceania, are on the right side of the verical line. As for Asia countries, they scatter abroad. So we can see that 
        1.Africa countries are different from countries of Europe, Americas and Oceania
        2.Countries of Europe, Americas and Oceania are similar
        3.Asia countries are varied
The former plot shows similar infomation. First, Africa countries are almost seperated from countries from the three continents. Most of Africa countries have PC1 less than 0. Second, countries of the three continent are almost clustered together. Third, Asia countries cross abroad the Africa cluster and the three continent cluster.




